sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-01-24 doc/paper-jss/pcod.Rmd
Copy editing

We will fit a geostatistical spatiotemporal model with \pkg{sdmTMB} for the purposes of index standardization. We will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound. The density units are kg/km^2^. Here, X and Y are UTM zone 9 coordinates in kilometres. If our spatial coordinates were not already in an equidistant projection, we would transform them [e.g., using `sf::st_transform()`, @pebesma2018] to ensure that distance remains constant throughout the study region or using the built-in `sdmTMB::add_utm_columns()` function.

@@

```{r plot-all-effects, fig.cap="Model predictions based on all fixed effects and random effects.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, "exp(est)") +
  # ggtitle("Prediction (fixed effects + all random effects)") +
  scale_fill_viridis_c(trans = "log10") 
```

@@

```{r plot-fix-defects, fig.cap="Predictions based on only the non-random field elements.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, "exp(est_non_rf)") +
  # ggtitle("Prediction (fixed effects only)") +
  scale_fill_viridis_c(trans = "sqrt")
```

We can look at the spatial random effects that represent consistent deviations in space through time that are not accounted for by our fixed effects (Fig.\ \ref{fig:plot-spatial-effects}). In other words, these deviations represent consistent biotic and abiotic factors that are affecting biomass density but are not accounted for in the model.

```{r plot-spatial-effects, fig.cap="Spatial random effects only.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, "omega_s") +
  # ggtitle("Spatial random effects only") +
  scale_fill_gradient2()
```

@@

```{r plot-spatiotemporal-effects, fig.cap="Spatiotemporal random effects only.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, "epsilon_st") +
  # ggtitle("Spatiotemporal random effects only") +
  scale_fill_gradient2()
```

@@

The above example uses `bias_correct = TRUE` in `get_index()`. The `bias_correct` argument is a feature of TMB for models with random effects, and more details can be found in @thorson2016bias. Leaving `bias_correct = FALSE` is fine for experimentation, but will be biased due to the nonlinear (`exp()`) transformation of the random effects. Setting `bias_correct = TRUE` implements a generic bias correction algorithm that should correct for this, but takes more time and memory than skipping the bias correction.

>>>>

We will fit a geostatistical spatiotemporal model with \pkg{sdmTMB} for the purposes of area-weighted biomass index standardization. We will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound. The density units are kg/km^2^. Here, X and Y are UTM zone 9 coordinates in kilometers. If our spatial coordinates were not already in an equidistant projection, we would transform them [e.g., using `sf::st_transform()`, @pebesma2018] to ensure that distance remains constant throughout the study region or using the built-in `sdmTMB::add_utm_columns()` function.

@@

```{r plot-all-effects, fig.cap="Model predictions based on all fixed effects and random effects.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, "exp(est)") +
  scale_fill_viridis_c(trans = "log10") 
```

@@

```{r plot-fix-defects, fig.cap="Predictions based on only the non-random field elements.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, "exp(est_non_rf)") +
  scale_fill_viridis_c(trans = "sqrt")
```

We can look at the spatial random effects that represent consistent deviations in space through time that are not accounted for by our fixed effects (Fig.\ \ref{fig:plot-spatial-effects}). In other words, these deviations represent consistent biotic and abiotic factors that are affecting spatial correlation in biomass density but are not accounted for in the model.

```{r plot-spatial-effects, fig.cap="Spatial random effects only.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, "omega_s") +
  scale_fill_gradient2()
```

@@

```{r plot-spatiotemporal-effects, fig.cap="Spatiotemporal random effects only.", fig.pos='ht', fig.align='center'}
plot_map(pred_qcs, "epsilon_st") +
  scale_fill_gradient2()
```

@@

The above example uses `bias_correct = TRUE` in `get_index()`. The `bias_correct` argument is a feature of TMB for models with random effects, and more details can be found in @thorson2016bias. Leaving `bias_correct = FALSE` is fine for experimentation, but will be biased due to the nonlinear (`exp()`) transformation of the random effects. Setting `bias_correct = TRUE` implements a generic bias correction algorithm that should correct for this [@thorson2016bias], but takes more time and memory than skipping the bias correction.

### 2023-01-24 doc/paper-jss/timing.Rmd
Copy editing

Here we describe the methods underlying the speed testing in Fig.\ \ref{fig:timing}.

We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Fig.\ \ref{fig:mesh-vis-timing}).
We did this by setting the `max.edge` argument, which controls the largest allowed triangle edge length.
We tested values of `max.edge` of 0.06, 0.075, 0.1, 0.15, and 0.2.
In Figure 4, we report on the x-axis the number of mesh nodes ("knots") that result from each of these meshes.

@@

We conducted 50 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values (Fig.\ \ref{fig:timing}).
The models were fit on a 2022 14 inch M1 MacBook Pro with an Apple M1 Pro 10-core CPU and 32 GB of RAM in R 4.1.3 and the default Matrix library. The following package versions were used:

>>>>

Here we describe the methods underlying the speed testing in Fig.\ \ref{fig:timing}.
We generated a mesh that was consistent across simulated data sets for a given mesh resolution (Fig.\ \ref{fig:mesh-vis-timing}).
We did this by setting the `max.edge` argument, which controls the largest allowed triangle edge length.
We tested values of `max.edge` of 0.06, 0.075, 0.1, 0.15, and 0.2.
In Figure 4, we report on the x-axis the number of mesh nodes ("knots" or "vertices") that result from each of these meshes.

@@

We conducted 50 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values (Fig.\ \ref{fig:timing}).
The models were fit on a 2022 14 inch M1 MacBook Pro with an Apple M1 Pro 10-core CPU and 32 GB of RAM in R 4.2.2 and the default Matrix library. The following package versions were used:
